---
title: "Reproducible Research: Course Project 1"
output: 
  html_document:
    keep_md: true
---

### Attach packages needed in the project
```{r,message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
```


### Loading and preprocessing the data
```{r}
unzip('activity.zip')
activity <- read.csv("./activity.csv")
activity$date <- as.Date(activity$date)
```


### What is mean total number of steps taken per day?

##### 1. Calculate the total number of steps taken per day
```{r}
steps_data <- activity %>% select(date, steps) %>% group_by(date) %>% summarize(StepsPerDay= sum(steps, na.rm = TRUE))
head(steps_data,10)
```

##### 2. Make a histogram of the total number of steps taken each day
```{r}
hist(steps_data$StepsPerDay, xlab = "Number of steps",main="Histogram of Total Steps taken per day", breaks = 10)
```

##### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
StepsPerDayMean <- mean(steps_data$StepsPerDay)
StepsPerDayMedian <- median(steps_data$StepsPerDay)
```
* Mean: `r StepsPerDayMean` 
* Median: `r StepsPerDayMedian`


## What is the average daily activity pattern?
##### 1. The time series plot
```{r}
interval_data <- activity %>% select(interval, steps) %>% group_by(interval) %>% summarize(AverageSteps = mean(steps, na.rm = TRUE))

ggplot(interval_data, aes(x=interval, y=AverageSteps)) + 
  geom_line() +
  labs(title="Steps By Time Interval", x="5 Minutes Intervals", y="Average Number of Steps") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

##### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
IntAvg <- round(max(interval_data$AverageSteps),4)
int <- interval_data$interval[which.max(interval_data$AverageSteps)]
```

* Interval `r int` contains the maximum number of steps with the average of `r IntAvg` steps.


### Imputing missing values

##### 1. Calculate and report the total number of missing values in the dataset
```{r}
NACounts <- sum(is.na(activity$steps))
```

* Number of missing values: `r NACounts`

##### 2. Devise a strategy for filling in all of the missing values in the dataset.
I will use the mean for that 5-minute interval to impute the missing values. 

##### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activityImputed <- activity %>% 
  group_by(interval) %>% 
  mutate(steps=ifelse(is.na(steps),mean(steps,na.rm=TRUE),steps)) %>% ungroup()
```


##### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
```{r}
steps_data_imputed <- activityImputed %>% select(date, steps) %>% group_by(date) %>% summarize(StepsPerDay= sum(steps, na.rm = TRUE))
head(steps_data_imputed)
```

```{r}
hist(steps_data_imputed$StepsPerDay, xlab = "Number of steps",main="Histogram of Total Steps taken per day (Imputed)", breaks = 10)
```

```{r}
StepsPerDayMeanImputed <- mean(steps_data_imputed$StepsPerDay)
StepsPerDayMedianImputed <- median(steps_data_imputed$StepsPerDay)
```
* Mean: `r StepsPerDayMeanImputed` 
* Median: `r StepsPerDayMedianImputed`

### Are there differences in activity patterns between weekdays and weekends?

##### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
activityImputed$weekday <- weekdays(activityImputed$date)
activityImputed$dayType <- ifelse(activityImputed$weekday=="Saturday" | activityImputed$weekday=="Sunday", "Weekend", "Weekday" )
```

##### 2. Make a panel plot containing a time series plot
```{r}
AggActivity <- aggregate(steps ~ interval + dayType, data=activityImputed, mean)

ggplot(AggActivity, aes(x=interval, y=steps, color=dayType)) + 
  geom_line() + 
  facet_grid(dayType ~ .) +
  labs(title="Steps By Time Interval - Weekday vs. Weekend", x="5 Minutes Intervals", y="Average Number of Steps") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.title = element_blank())
```
<br/> From the plot we can see the average steps fluctuate less on weekends while on weekdays there is a large spike in the morning. This is reasonable since most people commute on weekdays and don't need to go to work on the weekend.